Imperative Programming with Variational Effects

### 2018-04-03 paper/paper.tex
add motivation section for exceptions

\usepackage{setspace} 
\singlespacing
\usepackage{graphicx}                                        
\usepackage{amssymb}                                         
\usepackage{amsmath}                                         
\usepackage{amsthm}
\usepackage{url}         
\usepackage{tabu}

@@

>>>>

\usepackage{setspace} 
\singlespacing
\usepackage{graphicx}                                        
\usepackage{amssymb}                                         
\usepackage{amsmath}                                         
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{algorithmic}
\usepackage{url}         
\usepackage{tabu}

@@

\subsection{Exceptions}

Exceptions are a common type of effect that proves challenging in a variational setting.
Consider the following program:

\begin{algorithmic}
\STATE \ldots
\IF{$y$ is true}
\STATE{\textbf{throw} $e$}
\ENDIF
\STATE{$x \coloneqq$ myFunc()}
\end{algorithmic}

In a non-variational context, the behavior of this program should be clear.
If the variable $y$ evaluates to \textbf{true}, then an error is thrown with value
$e$. At this point evaluation should stop, meaning that the variable $x$ is never
assigned the value 2 in the final statement. This effect of halting execution and
returning an error value is the essence of exceptions in non-variational settings.

Now we consider the behavior of the same program in a variational context.
Suppose that the variable $y$ evaluates to the variational value
$A \langle \textbf{true}, \textbf{false} \rangle$. This means that in variant $A$ we
would evaluate the body of the if statement, while ignoring it otherwise. Therefore,
under variant $A$ we throw an exception, but otherwise we continue on to the next
final statement.

Clearly maintaining the same behavior from the non-variational setting in the variational setting
results in problems. If we halt execution whenever an exception is
thrown in any variant, then we stop the evaluation of variants that never encountered an error,
as in our example for the variant $\neg A$.

We also can't simply continue evaluation
in every variant regardless of whether or not we have encountered an error. This would be
a clear violation of the behavior we expect from exceptions.
Suppose that calling myFunc normally would evaluate to the variational value $A\langle 1, 2 \rangle$.
Additionally, suppose that arriving at the value 1 in variant $A$ is a fairly expensive action.
Because we know that variant $A$ will ultimately evaluate to the thrown exception $e$, we would
like to avoid this cost, mirroring how throwing an exception short-circuits evaluation in the
non-variational setting.

Another problem concerns how to keep track of which variants are in error states and what
the error values are. If we want to avoid the cost of pointless evaluation in variants that are in an error
state, we must have some efficient way of determining when evaluation is about to enter such a variant.
In the non-variational context we have no need to store and remember error values during evaluation
because we simply return the error value immediately when it is thrown. In the variational context, we
must now store these values while we continue to evaluate variants that are not in an error state.
